A scenario for a Planckian standard kilogram

### 2017-10-21 manuscript_gravitationalconstant.tex
Section 3 modified

In Fig.~\ref{fig:Gmeasurements} we have plotted some of the most recent measurements of the gravitational constant. 3 of them have been carried out using a torsion balance (), Schlamminger et al. used a beam balance and ....... measured the separation of two simple pendulums in a gravitational field. They all tried to measure the very weak gravitational force with a mechanical apparatus. The sixth measurement was performed under another approach, using cold atom interferometry. Its uncertainty is a lot larger than the others, but it could become more competitive in the future. However, a brief glance on the plotted measurements shows the disagreement of current $G$-values and suggests a demand for new, potentially different techniques.\\

@@

Now we are going to simulate position data of the two balls that the laser interferometry measurement would provide. From this data we then try to recover the parameters in the orbital equation of motion and thus also the gravitational constant. But first we have to find the orbital equation of motion for the two balls.\\

\subsection{Orbital equation of motion}
From Newtonian dynamics the equation of motion for a two-body system is given by 
\begin{equation}
\label{eq:eomeasy}
\ddot{\vec{r}} = - \frac{G M\, \vec{r}}{r^3}.
\end{equation}

In choosing the reference frame such that the net force in the system barycenter is zero, we can correct with a term in the equations of motion for the two-body system, which is proportional to the distance. Because we are in a 3-dimensional space, and possibly the influences of the different axis could mix, we have to multiply the distance vector $\vec{r}$ with a symmetric $3 \times 3$ matrix $C$, such that  
\begin{equation}
\label{eq:tidal}
\ddot{\vec{r}}_{\mathrm{C}} = C \cdot \vec{r}.
\end{equation}
or in terms of the individual components 
\begin{equation}
\label{eq:tidacom}
\left(\ddot{r}_\mathrm{C}\right)_i = \sum_{k=1}^{3} C_{ik} \: r_k.
\end{equation}
The matrix $C$ has six individual components which are not exactly known presumably, we will set constraints to their values and then fit for them. \\
Here $M$ denotes the sum of the masses $m_1$ and $m_2$, $G$ is the gravitational constant and $r$ is the relative distance between the two bodies. This equation just describes the motion of the two-body system. There are no forces acting on the center of mass, so the whole system is moving on a geodesic together with the spacecraft. For simplicity we will abbreviate the product of $GM$ with $k$.\\

Since our two-body system is in a rotating frame and there are perhaps other perturbations, we have to add the terms developed before for the Coriolis effect and the other corrections. Thus, the equation of motion for the two balls looks as follows:
\begin{equation}
\label{eq:eom}
\ddot{\vec{r}} = - \frac{k\,\vec{r}}{r^3} - 2 \; \vec{\omega} \times \dot{\vec{r}} + C \cdot \vec{r}
\end{equation}
These are actually three coupled equations of motion, one for each coordinate axis. Since all are second order differential equations, this leaves us with six initial conditions. In other words to determine the orbit, we need to give six starting values.\\
We choose our coordinate system such that the orbital plane of the two-body system is in the $x-y$ plane. $\vec{\omega}$ is a constant vector that points in the direction of the rotation axis of the earth-sun reference frame. We don't know the exact respective orientation of these two planes, but this is no problem, since we will fit for them. We just need to know the orientation to some degree, to provide good starting points for the fit.
The symmetric matrix $C$ has six independent components which account for tidal forces, centrifugal forces and other perturbations. Also here we don't have exact knowledge on how big these components will be and so we have to make an estimation and then fit for them. \\

Counting all unknown parameters in the equation of motion, we get 16. They correspond to 6 initial conditions, 6 components of the matrix $C$, 3 components of the vector $\vec{\omega}$ and the object of main interest, $k$,  which contains the gravitational constant ($k = GM$). Fitting for so many parameters is not an easy task, but it should be possible because the parameters are not completely unknown and thus we have good starting points for the fit.\\

@@

A Markov chain Monte-Carlo method is a Bayesian approach to data analysis. It uses the information provided by observed data about a set of parameters, to update a prior state of beliefs about the set of parameters to become a posterior state of beliefs about a set of parameters. First we have to specify the prior probability distribution, so to say what values for the 16 parameters in the equation of motion actually make sense. In table~\ref{tab:parameters}, estimated ranges for each parameters are shown. Then, for all 16 parameters a uniform distribution within these ranges and zero probability everywhere else is assumed as the prior probability distribution. \\

We start with the parameter $k$, which is the product of the gravitational constant $G$ and the total mass $M$. If we look at the measured values for the gravitational constant in Fig.~\ref{fig:Gmeasurements}, it is reasonable to constrain its numerical value in SI units to values between $6.671 \cdot 10^{-11}$ and $6.676 \cdot 10^{-11}$. The masses of the two balls can be determined to more than $10^{-5}$ in relative uncertainty, so they will not contribute to a larger uncertainty in $k$. If we take the consideration of balls with a mass of 6 kg, we can give a lower and an upper bound for $k$ which are displayed in table~\ref{tab:parameters}.\\

The vector $\vec{\omega}$ accounts for the rotation of the reference frame. If we say that the orbit of the balls lies in the same plane as the orbital plane of the earth, $\vec{\omega}$ will point in $z$ - direction with $x$ and $y$ components small. Thus $\omega_z$ is associated with the period $P_\odot$ of one year through $2\pi/P_\odot$. Because the rotation axis are conceivably not exactly aligned, we add some uncertainty. As mentioned before, the $x$ and $y$ components are small, so to say roughly in the range of the uncertainty in $w_z$.\\

To estimate the six components of the matrix $C$, we first state again that the barycentre of the two balls is in free fall. So at this point all forces cancel out. If we imagine the ball on its orbit, the tidal force is stronger when its nearer to the earth-sun barycentre and the centrifugal force is stronger when it is in the opposite position. So over an orbit, the effect on the balls stays constant. Also other perturbations are assumed to be constant, at least during some orbits.
The main component of this term should be perpendicular to the rotation axis, for which we have chosen the $z$ - axis. Since we are free to set the $x$ and $y$ axis in the orbital plane, we can choose it such that the main contribution lies in the $x$ - axis. Its magnitude goes as $\omega^2$, so the $C_{xx}$ component should be around $4 \cdot 10^{-14}$. Because we assumed that the main effect is in the $x$ - direction, all the other components are smaller, we constrain them to somewhere between $\pm 10^{-14}$. Since the $C$ matrix also absorbs other perturbations, we give some more uncertainty here. \\

The initial values for the positions can be guessed very accurately, since the position is directly measured precisely. We choose the coordinate system such that the orbit is in the $x-y$ plane and the origin is at the barycenter of the two balls. We set the $x$ - axis such that at the beginning it corresponds to the radius and the $y$ - component is zero.
If we choose the initial positions like this, the initial velocities in $x-$ and $z$ - direction are very small, they just arise from perturbations. The velocity in $y$ - direction can be calculated from Kepler's third law and is therefore also linked with the parameter $k$.\\

@@

To simulate data we would obtain from the position measurements, we integrate the equations of motion. The time for one orbit without perturbations can be estimated using Kepler's third law (\ref{eq:kepler3}). For the value of $k$ assumed above and a semi-major axis of $0.15 \, \mathrm{m}$, this results in an orbital period of roughly $12900 \, \mathrm{seconds}$ or $3.5 \, \mathrm{hours}$. Here we integrate two orbits in $100000$ steps, that would correspond to about $4$ measurements per second, which should be feasible. To the generated data, we add normal distributed noise with a standard deviation of $10 \, \mathrm{nm}$. This uncertainty stays for a systematic error in the laser interferometry measurement and also random statistic noise. Its value is just an estimation, it would depend highly on the measurement system that has been used. There are laser interferometers, which are able to measure positions with an accuracy of $10\,\mathrm{nm}$.\cite{Loughridge13} \\

A useful way to look at the samples of the posterior returned by the MCMC is doing a corner plot using a Python module named \textit{corner}\cite{corner}. The corner plot shows all the one and two dimensional projections of the posterior probability distributions and demonstrates quickly the covariances between parameters. Along the diagonal, the marginalized distribution for each parameter independently is shown. From this distribution we can then estimate its best value and uncertainty.
In Fig.~\ref{fig:plotCorner4} an excerpt of the full corner plot is shown. Here we only depict the parameters $x_0$, $v_{y_0}$, $\omega_z$, $C_{xx}$ and $k$. These are the more important ones, because for example the main contribution to $\vec{\omega}$ comes from its $z$-component.\\

@@

To arrive at a best value for $k$ and an uncertainty, we take the median and the percentiles corresponding to one sigma from this distribution of $k$. For this example we get:
\begin{equation}\label{eq:kfit}
k = \left( 8.0088006 \, \substack{+0.0000005 \\ -0.0000008}\, \right) \cdot 10^{-10}
\end{equation}
Now we divide $k$ by the total mass of the two balls. Note, that we assume, that the uncertainty on the mass of the balls does not contribute to a larger uncertainty on $G$, because their mass should be determinable to a relative uncertainty of about $10^{-8}$, whereas the one of $k$ is roughly $10^{-7}$. We change the way to write the uncertainty from the 0.16 and 0.84 percentiles to the usual standard deviation, while adding some margin. With units put back in, we arrive at the following value for the gravitational constant $G$:
\begin{equation}\label{Gfit}
G = \left(6.674005 \pm 0.000001 \right) \cdot 10^{-11}  \frac{\mathrm{m}^3}{\mathrm{kg \, s}^2}
\end{equation}
This corresponds to a relative uncertainty of $1.5 \cdot 10^{-7}$ or 15 parts in a hundred million.\\

@@

>>>>

In Fig.~\ref{fig:Gmeasurements} we have plotted some of the most recent measurements of the gravitational constant. Three of them have been carried out using a torsion balance (BIPM-14\cite{Quinn14}, UCI-14\cite{Newman14}, HUST-09\cite{Luo09}), UZur-06\cite{Schlamminger06} used a beam balance and JILA-10\cite{Parks10} measured the separation of two simple pendulums in a gravitational field. They all tried to measure the very weak gravitational force with a mechanical apparatus. The sixth measurement (LENS-14\cite{Rosi14}) was performed under another approach, using cold atom interferometry. Its uncertainty is a lot larger than the others, but it could become more competitive in the future. However, a brief glance on the plotted measurements shows the disagreement of current $G$-values and suggests a demand for new, potentially different techniques.\\

@@

In this section we are going to show how the gravitational constant can be recovered from position data of the two balls that the laser interferometry measurement would provide.
But first we have to find the orbital equation of motion for the two balls.\\

\subsection{Orbital equation of motion}
From Newtonian dynamics the equation of motion for a clean two-body system is given by 
\begin{equation}
\label{eq:eomeasy}
\ddot{\vec{r}} = - \frac{G M\, \vec{r}}{r^3}.
\end{equation}
Here $M$ denotes the sum of the masses $m_1$ and $m_2$, $G$ is the gravitational constant and $r$ is the relative distance between the two bodies. For simplicity we will abbreviate the product of $GM$ with $k$.
Since out two-body system is in a rotating frame we have to add the term developed before for the Coriolis acceleration as well as a correction term proportional to the distance accounting for centrifugal and tidal forces. This last term can be written with a
symmetric $3 \times 3$ matrix $C$ as follows  
\begin{equation}
\label{eq:tidal}
\ddot{\vec{r}}_{\mathrm{C}} = C \cdot \vec{r}.
\end{equation}
or in terms of the individual components 
\begin{equation}
\label{eq:tidacom}
\left(\ddot{r}_\mathrm{C}\right)_i = \sum_{k=1}^{3} C_{ik} \: r_k.
\end{equation}
It can be thought of as a first order expansion and therefore also include other possible minor perturbations. The reference frame is chosen such that the barycenter of the two-body system is force-free.
So the full equation of motion for the two balls looks as follows:
\begin{equation}
\label{eq:eom}
\ddot{\vec{r}} = - \frac{k\,\vec{r}}{r^3} - 2 \; \vec{\omega} \times \dot{\vec{r}} + C \cdot \vec{r}
\end{equation}
These are actually three coupled equations of motion, one for each coordinate axis. Since all are second order differential equations, this leaves us with six initial conditions. In other words to determine the orbit, we need to give six starting values.\\
We choose our coordinate system such that the orbital plane of the two-body system is in the $x-y$ plane. $\vec{\omega}$ is a constant vector that points in the direction of the rotation axis of the earth-sun reference frame. We will fit for $\vec{\omega}$, so we don't have to know it exactly before, but it can be estimated well and thus provides good starting points for the fit. The same is valid for the six components of the matrix $C$.\\

Counting all unknown parameters in the equation of motion, we get 16. They correspond to 6 initial conditions, 6 components of the matrix $C$, 3 components of the vector $\vec{\omega}$ and the object of main interest, $k$,  which contains the gravitational constant ($k = GM$). Fitting for so many parameters is not an easy task, but it should be possible because the parameters are not completely unknown.\\

@@

A Markov chain Monte-Carlo method is a Bayesian approach to data analysis. It uses the information provided by observed data about a set of parameters, to update a prior state of beliefs about the set of parameters to become a posterior state of beliefs about a set of parameters. First we have to specify the prior probability distribution, so to say what values for the 16 parameters in the equation of motion actually make sense. Then, for all 16 parameters a uniform distribution within these ranges and zero probability everywhere else is assumed as the prior probability distribution. \\

We briefly explain in what range the values of the parameters are estimated and display it in table~\ref{tab:parameters}.
For the parameter $k$, which is the product of the gravitational constant $G$ and the total mass $M$ we look at the current measured values for the gravitational constant in Fig.~\ref{fig:Gmeasurements}, it is reasonable to constrain its numerical value in SI units to values between $6.671 \cdot 10^{-11}$ and $6.676 \cdot 10^{-11}$. The masses of the two balls will not contribute to a larger uncertainty in $k$, we choose them to be 6 kg.\\
The vector $\vec{\omega}$ accounts for the rotation of the reference frame. If we say that the orbit of the balls lies in the same plane as the orbital plane of the earth, $\vec{\omega}$ will point in $z$ - direction with $x$ and $y$ components small. Thus $\omega_z$ is associated with the period $P_\odot$ of one year through $2\pi/P_\odot$. Because the rotation axis are conceivably not exactly aligned, we add some uncertainty. As mentioned before, the $x$ and $y$ components are small, so to say roughly in the range of the uncertainty in $w_z$.\\
The main contribution to the tidal term goes with $\omega^2$ and lies in the orbital plane. We are free to choose the $x$ and $y$ axis, so we choose it such that it is in $x$-direction. The other five components then should be smaller, but we give some extra range to account for possible other perturbations.\\
% 
%To estimate the six components of the matrix $C$, we first state again that the barycentre of the two balls is in free fall. So at this point all forces cancel out. If we imagine the ball on its orbit, the tidal force is stronger when its nearer to the earth-sun barycentre and the centrifugal force is stronger when it is in the opposite position. So over an orbit, the effect on the balls stays constant. Also other perturbations are assumed to be constant, at least during some orbits.
%The main component of this term should be perpendicular to the rotation axis, for which we have chosen the $z$ - axis. Since we are free to set the $x$ and $y$ axis in the orbital plane, we can choose it such that the main contribution lies in the $x$ - axis. Its magnitude goes as $\omega^2$, so the $C_{xx}$ component should be around $4 \cdot 10^{-14}$. Because we assumed that the main effect is in the $x$ - direction, all the other components are smaller, we constrain them to somewhere between $\pm 10^{-14}$. Since the $C$ matrix also absorbs other perturbations, we give some more uncertainty here. \\
%
The initial values for the positions can be guessed very accurately, since they are directly measured precisely. The initial velocity can be calculated easily and matched to the initial positions, therefore also yielding narrow parameter ranges.\\

@@

To simulate data we would obtain from the position measurements, we integrate the equations of motion. The time for one orbit without perturbations can be estimated using Kepler's third law (\ref{eq:kepler3}). For the value of $k$ assumed above and a semi-major axis of $0.15 \, \mathrm{m}$, this results in an orbital period of roughly $12900 \, \mathrm{seconds}$ or $3.5 \, \mathrm{hours}$. Here we integrate two orbits in $100000$ steps, that would correspond to about $4$ measurements per second, which should be feasible. To the generated data, we add normal distributed noise with a standard deviation of $10 \, \mathrm{nm}$. This uncertainty stays for a systematic error in the laser interferometry measurement and also random statistic noise. Its value is just an estimation, it would depend highly on the measurement system that has been used.\\

A useful way to look at the samples of the posterior returned by the MCMC is doing a corner plot using a Python module named \textit{corner}\cite{corner}. The corner plot shows all the one and two dimensional projections of the posterior probability distributions and demonstrates quickly the covariances between parameters. Along the diagonal, the marginalized distribution for each parameter independently is shown.
In Fig.~\ref{fig:plotCorner4} an excerpt of the full corner plot is shown. Here we only depict the parameters $x_0$, $v_{y_0}$, $\omega_z$, $C_{xx}$ and $k$. These are the more important ones, because for example the main contribution to $\vec{\omega}$ comes from its $z$-component.\\

@@

To arrive at a best value for $k$ and an uncertainty, we take the median and the percentiles corresponding to one sigma from the distribution of the parameter $k$. For this example we get:
\begin{equation}\label{eq:kfit}
k = \left( 8.0088006 \, \substack{+0.0000005 \\ -0.0000008}\, \right) \cdot 10^{-10}
\end{equation}
Now we divide $k$ by the total mass of the two balls. Note, that we assume, that the uncertainty on the mass of the balls does not contribute to a larger uncertainty on $G$, because their mass should be determinable to a relative uncertainty of about $10^{-8}$, whereas the one of $k$ is roughly $10^{-7}$. We change the way to write the uncertainty from the 0.16 and 0.84 percentiles to the usual standard deviation, while adding some margin. With units put back in, we arrive at the following value for the gravitational constant $G$:
\begin{equation}\label{Gfit}
G = \left(6.674005 \pm 0.000001 \right) \cdot 10^{-11}  \frac{\mathrm{m}^3}{\mathrm{kg \, s}^2}
\end{equation}
This corresponds to a relative uncertainty of $1.5 \cdot 10^{-7}$ or 15 parts in a hundred million.\\

@@

\bibitem{Quinn14} T. Quinn, C. Speake, H. Parks and R. Davis, ``The BIPM measurements of the Newtonian constant of gravitation $G$'', \textit{Philosophical Transactions of the Royal Society of London A: Mathematical, Physical and Engineering Sciences}, 372(2026) (September 2014)

\bibitem{Newman14} R. Newman, M. Bantel, E. Berg and W. Cross, ``A measurement of $G$ with a cryogenic torsion pendulum'', \textit{Philosophical Transactions of the Royal Society of London A: Mathematical, Physical and Engineering Sciences}, 372(2026) (September 2014)

\bibitem{Luo09} J. Luo, W. Liu, L. Tu, C. Shao, L.Liu, S. Yang, Q. Li and Y. Zhang, ``Determination of the Newtonian gravitational constant $G$ with time-of-swing method'', \textit{Phys. Rev. Lett.}, 102:240801 (June 2009)

\bibitem{Schlamminger06} S. Schlamminger, E. Holzschuh, W. Kündig, F. Nolting, R.E. Pixley, J. Schnurr and U. Straumann, ``Measurement of Newton's gravitational constant", \textit{Phys. Rev. D}, 74:082001 (October 2009)

\bibitem{Parks10} H.V. Parks and J.E. Faller, ``Simple pendulum determination of the gravitational constant", \textit{Phys. Rev. Lett.}, 105:110801 (September 2010)

\bibitem{Rosi14} G. Rosi, F. Sorrentino, L. Cacciapuoti, M. Prevedelli and G.M. Tino, ``Precision measurement of the Newtonian gravitational constant using cold atoms", \textit{Nature}, 510:518-521 (June 2014)
